Near Optimal Behavior via Approximate State Abstraction

### 2016-02-05 paper/results_discussion.tex
Added justification for just doing Q* and not other abstractions in empirical results as well as replaced a bunch of Q values with Q-values.

\section{Approximate Abstraction on Example Domains}
%Explanation of what abstraction is used
We ran experiments on aggregation functions in the $\epQ$ family of approximate abstractions.  \dnote{Some explanation of why we didn't do all 4 abstractions}
For the purpose of illustrating what kinds of approximations are possible, we built each abstraction by first solving the MDP, then greedily aggregating ground states into abstract states that satisfied the $\epQ$ criteria. Since this approach represents an order-dependent approximation to the maximum amount of abstraction possible, we randomized the order in which states were considered across trials. Every ground state is equally weighted in its abstract state.

>>>>

\section{Approximate Abstraction on Example Domains}
%Explanation of what abstraction is used
We ran experiments on aggregation functions in the $\epQ$ family of approximate abstractions. We provide results for only $\epQ$ because, as our proofs in Section \ref{sec:theory_results} demonstrate, the other 3 functions are reducable to particular $\epQ$ functions. For the purpose of illustrating what kinds of approximations are possible, we built each abstraction by first solving the MDP, then greedily aggregating ground states into abstract states that satisfied the $\epQ$ criteria. Since this approach represents an order-dependent approximation to the maximum amount of abstraction possible, we randomized the order in which states were considered across trials. Every ground state is equally weighted in its abstract state.

### 2016-02-05 paper/theory_results.tex
Added justification for just doing Q* and not other abstractions in empirical results as well as replaced a bunch of Q values with Q-values.

% Intro to approx state abstraction results.
Here, we introduce our formal analysis of approximate state abstraction, including results bounding the error associated with these abstraction methods. In particular, we demonstrate that abstractions based on approximate $Q^*$ similarity (\ref{sec:Q*}), approximate model similarity (\ref{sec:model}), and approximate similarity between distributions over $Q^*$, for both Boltzmann (\ref{sec:boltz}) and multinomial (\ref{sec:mult}) distributions, induce abstract \acp{MDP} for which the optimal policy has bounded error in the ground MDP.

@@

% Proof
\textbf{Proof of Lemma~\ref{lma:Q*}:}
We first demonstrate that $Q$-values in the abstract \ac{MDP} are boundedly close to $Q$-values in the ground \ac{MDP} (Claim \ref{clm:closeQs}). We next leverage Claim \ref{clm:closeQs} to demonstrate that the optimal action in the abstract \ac{MDP} is nearly optimal in the ground \ac{MDP}  (Claim \ref{clm:optAbsActionNearOptGround}). Lastly, we use Claim \ref{clm:optAbsActionNearOptGround} to conclude Lemma \ref{lma:Q*} (Claim \ref{clm:lmaFromClm}).
% Claim 1:
\begin{clm}
\label{clm:closeQs}
Optimal $Q$ values in the abstract \ac{MDP} closely resemble optimal $Q$ values in the ground \ac{MDP}:
\begin{equation}
\label{eq:Q*Claim1}
\forall_{s_G \in \mathcal{S}_G, a} |Q_G(s_G, a) - Q_A(\epQ(s_G), a)| \leq \frac{\epsilon}{1-\gamma}.
\end{equation}
\end{clm}

@@

It follows that the $Q$ value of a state $s$ in $M^T$ for action $a$ is defined by:
\begin{equation}
Q_T(s, a) = 
\begin{cases}
	   Q_G(s, a), &  \text{if } T=0,\\
	   \underset{g \in G(s)}{\sum} \left[ \omega(g)Q_G(g,a) \right], & \text{if } T = 1,\\
	   \mathcal{R}_A(\epQ(s),a) + \sigma_{T-1}(s,a), &\text{o/w},
\end{cases}
\end{equation}
where
\begin{equation}
\sigma_{T-1}(s,a) = \gamma \underset{{s_A}' \in \mathcal{S}_A}{\sum} \mathcal{T}_A(\epQ(s),a,{s_A}') \max_{a'} Q_{T-1}({s_A}', a').
\end{equation}

@@

% Base Case t=1
\textit{Base Case: $T = 1$}
\begin{align*}
&Q_1(s,a) = \underset{g \in G(s)}{\sum} \left[ \omega(g)Q_G(g,a) \right].
\end{align*}
Since all co-aggregated states have $Q$ values within $\epsilon$ of one another and $\omega(g)$ induces a convex combination,
\begin{align*}
&Q_1(s,a) \leq \epsilon \gamma^t + \epsilon + Q_G(s_G, a).
\end{align*}
Thus,
\begin{equation}
\left| Q_{1}(s_A, a) - Q_G(s_G,a) \right| \leq \sum_{t=0}^{1}\epsilon \gamma^t.
\end{equation}
% Inductive Case
\textit{Inductive Case: $T > 1$}

@@

Consider a fixed but arbitrary state, $s_G \in \mathcal{S}_G$, and fixed but arbitrary action $a$.
We denote $\epQ(s_G)$ as $s_A$. 
By definition of $Q_{T}(s_A, a)$, $\mathcal{R}_A$, $\mathcal{T}_A$:
\begin{multline*}
Q_T(s_A, a) = \sum_{g \in G(s_A)}\omega(g)\ \times \\ 
 \left[ R_0(g,a) + \gamma \sum_{g' \in \mathcal{S}_G} T_0(g,a,g') \max_{a'} Q_{T-1}(g', a')      \right].
\end{multline*}
Applying our inductive hypothesis yields:
\begin{multline*}
Q_T(s_A, a) \leq \sum_{g \in G(s_A)}\omega(g) \times \biggl[ R_G(g,a)\ + \\ \gamma \sum_{g' \in \mathcal{S}_G} T_G(g,a,g') \max_{a'}(Q_G(g', a') + \sum_{t=0}^{T-2} \epsilon \gamma^t) \biggr].
\end{multline*}
%&\leq \gamma\sum_{t=0}^{T-2} \epsilon \gamma^t + \sum_{g \in X(s_A)}\omega(g)\left[ R_0(g,a) + \gamma \sum_{g' \in \mathcal{S}_G} T_0(g,a,g') \max_{a'}Q_0      \right]\\
Then,
\begin{equation*}
Q_T(s_A, a) \leq \gamma\sum_{t=0}^{T-2} \epsilon \gamma^t + \sum_{g \in G(s_A)}\left[ \omega(g)\ Q_G(g,a)\right].
\end{equation*}
since all aggregated states have $Q$ values within $\epsilon$ of one another:
\begin{align*}
Q_T(s_A, a) \leq \gamma\sum_{t=0}^{T-2} \epsilon \gamma^t + \epsilon + Q_G(s_G, a). \\
\therefore\ \left| Q_{T}(s_A, a) - Q_G(s_G,a) \right| \leq \gamma\sum_{t=0}^{T-1}\epsilon \gamma^t
\end{align*}
Since $s_G$ and $s_A$ are arbitrary, we conclude
\begin{equation}
\forall_{T, s_G \in \mathcal{S}_G, a} |Q_G(s_G, a) - Q_T(\epQ(s_G), a)| \leq \sum_{t=0}^{T-1} \epsilon \gamma^t.
\end{equation}

@@

%Claim 2
\begin{clm}
\label{clm:optAbsActionNearOptGround}
%The optimal action in the abstract MDP has a $Q$ value in the ground which is nearly optimal:

Consider a fixed but arbitrary state, $s_G \in \mathcal{S}_G$ and its corresponding abstract state $s_A=\epQ(s_G)$.
Let $a^*_G$ stand for the optimal action in $s_G$, and $a^*_A$ stand for the optimal action in $s_A$:
\begin{align*}
a^*_G = \argmin Q_G(s_G, a), \hspace{4mm}
a^*_A = \argmin Q_A(s_A, a).
\end{align*}
The optimal action in the abstract MDP has a $Q$ value in the ground MDP that is nearly optimal:
\begin{equation}
\label{eq:Q*Claim2}
V_G(s_G) \leq Q_G(s_A, a^*_A) + \frac{2\epsilon}{1-\gamma}.
\end{equation}
\end{clm}

@@

{\bf Proof of Lemma~\ref{lma:model}:} Our proof proceeds as follows. We demonstrate that states aggregated under $\epM$ have $Q$ values all within a particular bound. Consequently, $\epM$ abstractions are interpretable as an instantiation of $\epQ$ with a particular $\epsilon$. Understanding $\epM$ as such an abstraction allows us to conclude our bound by applying Lemma \ref{lma:Q*}.

Let $B$ stand for the maximum $Q$ value difference between any pair of ground states in the same abstract state under $\epM$:
\begin{equation*}
B = \max_{s_A, s_1, s_2, a}  |Q_G(s_1, a) - Q_G(s_2, a)|,
\end{equation*}
where $s_A \in \mcS_G$ and $s_1, s_2 \in G(s_A)$:
\begin{multline*}
B=\max_{s_A, s_1, s_2, a}      \biggl|\mcR_G(s_1, a) - \mcR_G(s_2, a)\ + \gamma \sum_{{s_G}' \in \mcS_G} \\
\biggl[(\mcT_G(s_1,a,{s_G}') -\mcT_G(s_2, a, {s_G}'))\max_{a'}Q_G({s_G}', a')\biggr]\biggr|
\end{multline*}
\begin{multline*}
B\leq \epsilon + \gamma \sum_{s_A \in \mcS_A}\sum_{{s_G}' \in G(s_A)} \biggl[(T_G(s_1, a, {s_G}')\ -\\ 
T_G(s_2, a, {s_G}')) \max_{a'}Q_G({s_G}', a') \biggr].
\end{multline*}
By similarity of transitions under $\epM$:
\begin{align*}
B \leq \epsilon + \gamma \textsc{QMax} \sum_{s_A \in \mcS_A} \epsilon \leq \epsilon + \gamma|\mcS_G|\epsilon \textsc{QMax}.
\end{align*}
Since \textsc{QMax} $\leq \frac{\textsc{RMax}}{1-\gamma}$, and we defined $\textsc{RMax} = 1$:
\begin{equation*}
B \leq \frac{\epsilon + \gamma(|\mcS_G| - 1) \epsilon}{1-\gamma}.
\end{equation*}
Since the $Q$ values of ground states grouped under $\epM$ are strictly less than $B$, we can understand $\epM$ as a type of $\epQ$ with $\epsilon = B$. Applying Lemma \ref{lma:Q*} yields Lemma \ref{lma:model}.
\subsection{Boltzmann over Optimal Q: $\epB$}
\label{sec:boltz}

%zzz all this theory is quite abstract. is there some way to weave in a running example and/or to at least provide some more motiviation for why phiBoltz is desirable. Balancing exploration/exploitation isn't really a motivation for why values should be abstracted together. What do the first two abstraction types miss that this one provides? We didn't Lihong et al. study it in the exact case?
Here, we introduce the approximate abstraction family, $\epB$, which aggregates states with similar Boltzmann distributions on optimal $Q$ values as equivalent. This family of abstractions is appealing as the Boltzmman distribution naturally balances exploration and exploitation~\cite{sutton1998reinforcement}.

@@

By the approximation in Equation~\ref{eq:e_to_x_approx} and the assumption in Equation~\ref{eq:bolt_denom}:
\begin{align*}
\left|\frac{1 + Q_G(s_1,a) + \delta_1}{\sum_j e^{Q_G(s_1,a_j)}} - \frac{1 + Q_G(s_2,a) + \delta_2}{\sum_j e^{Q_G(s_1,a_j)} \pm k\epsilon}\right| \leq \epsilon
\end{align*}
It follows that:
\begin{equation}
\left|Q_G(s_1,a) - Q_G(s_2,a)\right| \leq \epsilon \left(\frac{|\mathcal{A}|}{1-\gamma} + k\epsilon + k \right).
\label{eq:bolt_qs}
\end{equation}
Consequently, we can consider $\epB$ as a variation of $\epQ$, where the difference in $Q$ values is bounded by $\epsilon \left(\frac{|\mathcal{A}|}{1-\gamma} + k\epsilon + k \right)$. Lemma~\ref{lma:bolt_lemma} then follows directly from Lemma~\ref{lma:Q*}.

>>>>

\label{sec:theory_results}
% Intro to approx state abstraction results.
Here, we introduce our formal analysis of approximate state abstraction, including results bounding the error associated with these abstraction methods. In particular, we demonstrate that abstractions based on approximate $Q^*$ similarity (\ref{sec:Q*}), approximate model similarity (\ref{sec:model}), and approximate similarity between distributions over $Q^*$, for both Boltzmann (\ref{sec:boltz}) and multinomial (\ref{sec:mult}) distributions, induce abstract \acp{MDP} for which the optimal policy has bounded error in the ground MDP.

@@

% Proof
\textbf{Proof of Lemma~\ref{lma:Q*}:}
We first demonstrate that $Q$-values in the abstract \ac{MDP} are boundedly close to $Q$-values in the ground \ac{MDP} (Claim \ref{clm:closeQs}). We next leverage Claim \ref{clm:closeQs} to demonstrate that the optimal action in the abstract \ac{MDP} is nearly optimal in the ground \ac{MDP}  (Claim \ref{clm:optAbsActionNearOptGround}). Lastly, we use Claim \ref{clm:optAbsActionNearOptGround} to conclude Lemma \ref{lma:Q*} (Claim \ref{clm:lmaFromClm}).
% Claim 1:
\begin{clm}
\label{clm:closeQs}
Optimal $Q$-values in the abstract \ac{MDP} closely resemble optimal $Q$-values in the ground \ac{MDP}:
\begin{equation}
\label{eq:Q*Claim1}
\forall_{s_G \in \mathcal{S}_G, a} |Q_G(s_G, a) - Q_A(\epQ(s_G), a)| \leq \frac{\epsilon}{1-\gamma}.
\end{equation}
\end{clm}

@@

It follows that the $Q$-value of a state $s$ in $M^T$ for action $a$ is defined by:
\begin{equation}
Q_T(s, a) = 
\begin{cases}
	   Q_G(s, a), &  \text{if } T=0,\\
	   \underset{g \in G(s)}{\sum} \left[ \omega(g)Q_G(g,a) \right], & \text{if } T = 1,\\
	   \mathcal{R}_A(\epQ(s),a) + \sigma_{T-1}(s,a), &\text{o/w},
\end{cases}
\end{equation}
where
\begin{equation}
\sigma_{T-1}(s,a) = \gamma \underset{{s_A}' \in \mathcal{S}_A}{\sum} \mathcal{T}_A(\epQ(s),a,{s_A}') \max_{a'} Q_{T-1}({s_A}', a').
\end{equation}

@@

% Base Case t=1
\textit{Base Case: $T = 1$}
\begin{align*}
&Q_1(s,a) = \underset{g \in G(s)}{\sum} \left[ \omega(g)Q_G(g,a) \right].
\end{align*}
Since all co-aggregated states have $Q$-values within $\epsilon$ of one another and $\omega(g)$ induces a convex combination,
\begin{align*}
&Q_1(s,a) \leq \epsilon \gamma^t + \epsilon + Q_G(s_G, a).
\end{align*}
Thus,
\begin{equation}
\left| Q_{1}(s_A, a) - Q_G(s_G,a) \right| \leq \sum_{t=0}^{1}\epsilon \gamma^t.
\end{equation}
% Inductive Case
\textit{Inductive Case: $T > 1$}

@@

Consider a fixed but arbitrary state, $s_G \in \mathcal{S}_G$, and fixed but arbitrary action $a$.
We denote $\epQ(s_G)$ as $s_A$. 
By definition of $Q_{T}(s_A, a)$, $\mathcal{R}_A$, $\mathcal{T}_A$:
\begin{multline*}
Q_T(s_A, a) = \sum_{g \in G(s_A)}\omega(g)\ \times \\ 
 \left[ R_0(g,a) + \gamma \sum_{g' \in \mathcal{S}_G} T_0(g,a,g') \max_{a'} Q_{T-1}(g', a')      \right].
\end{multline*}
Applying our inductive hypothesis yields:
\begin{multline*}
Q_T(s_A, a) \leq \sum_{g \in G(s_A)}\omega(g) \times \biggl[ R_G(g,a)\ + \\ \gamma \sum_{g' \in \mathcal{S}_G} T_G(g,a,g') \max_{a'}(Q_G(g', a') + \sum_{t=0}^{T-2} \epsilon \gamma^t) \biggr].
\end{multline*}
%&\leq \gamma\sum_{t=0}^{T-2} \epsilon \gamma^t + \sum_{g \in X(s_A)}\omega(g)\left[ R_0(g,a) + \gamma \sum_{g' \in \mathcal{S}_G} T_0(g,a,g') \max_{a'}Q_0      \right]\\
Then,
\begin{equation*}
Q_T(s_A, a) \leq \gamma\sum_{t=0}^{T-2} \epsilon \gamma^t + \sum_{g \in G(s_A)}\left[ \omega(g)\ Q_G(g,a)\right].
\end{equation*}
since all aggregated states have $Q$-values within $\epsilon$ of one another:
\begin{align*}
Q_T(s_A, a) \leq \gamma\sum_{t=0}^{T-2} \epsilon \gamma^t + \epsilon + Q_G(s_G, a). \\
\therefore\ \left| Q_{T}(s_A, a) - Q_G(s_G,a) \right| \leq \gamma\sum_{t=0}^{T-1}\epsilon \gamma^t
\end{align*}
Since $s_G$ and $s_A$ are arbitrary, we conclude
\begin{equation}
\forall_{T, s_G \in \mathcal{S}_G, a} |Q_G(s_G, a) - Q_T(\epQ(s_G), a)| \leq \sum_{t=0}^{T-1} \epsilon \gamma^t.
\end{equation}

@@

%Claim 2
\begin{clm}
\label{clm:optAbsActionNearOptGround}
%The optimal action in the abstract MDP has a $Q$-value in the ground which is nearly optimal:

Consider a fixed but arbitrary state, $s_G \in \mathcal{S}_G$ and its corresponding abstract state $s_A=\epQ(s_G)$.
Let $a^*_G$ stand for the optimal action in $s_G$, and $a^*_A$ stand for the optimal action in $s_A$:
\begin{align*}
a^*_G = \argmin Q_G(s_G, a), \hspace{4mm}
a^*_A = \argmin Q_A(s_A, a).
\end{align*}
The optimal action in the abstract MDP has a $Q$-value in the ground MDP that is nearly optimal:
\begin{equation}
\label{eq:Q*Claim2}
V_G(s_G) \leq Q_G(s_A, a^*_A) + \frac{2\epsilon}{1-\gamma}.
\end{equation}
\end{clm}

@@

{\bf Proof of Lemma~\ref{lma:model}:} Our proof proceeds as follows. We demonstrate that states aggregated under $\epM$ have $Q$-values all within a particular bound. Consequently, $\epM$ abstractions are interpretable as an instantiation of $\epQ$ with a particular $\epsilon$. Understanding $\epM$ as such an abstraction allows us to conclude our bound by applying Lemma \ref{lma:Q*}.

Let $B$ stand for the maximum $Q$-value difference between any pair of ground states in the same abstract state under $\epM$:
\begin{equation*}
B = \max_{s_A, s_1, s_2, a}  |Q_G(s_1, a) - Q_G(s_2, a)|,
\end{equation*}
where $s_A \in \mcS_G$ and $s_1, s_2 \in G(s_A)$:
\begin{multline*}
B=\max_{s_A, s_1, s_2, a}      \biggl|\mcR_G(s_1, a) - \mcR_G(s_2, a)\ + \gamma \sum_{{s_G}' \in \mcS_G} \\
\biggl[(\mcT_G(s_1,a,{s_G}') -\mcT_G(s_2, a, {s_G}'))\max_{a'}Q_G({s_G}', a')\biggr]\biggr|
\end{multline*}
\begin{multline*}
B\leq \epsilon + \gamma \sum_{s_A \in \mcS_A}\sum_{{s_G}' \in G(s_A)} \biggl[(T_G(s_1, a, {s_G}')\ -\\ 
T_G(s_2, a, {s_G}')) \max_{a'}Q_G({s_G}', a') \biggr].
\end{multline*}
By similarity of transitions under $\epM$:
\begin{align*}
B \leq \epsilon + \gamma \textsc{QMax} \sum_{s_A \in \mcS_A} \epsilon \leq \epsilon + \gamma|\mcS_G|\epsilon \textsc{QMax}.
\end{align*}
Since \textsc{QMax} $\leq \frac{\textsc{RMax}}{1-\gamma}$, and we defined $\textsc{RMax} = 1$:
\begin{equation*}
B \leq \frac{\epsilon + \gamma(|\mcS_G| - 1) \epsilon}{1-\gamma}.
\end{equation*}
Since the $Q$-values of ground states grouped under $\epM$ are strictly less than $B$, we can understand $\epM$ as a type of $\epQ$ with $\epsilon = B$. Applying Lemma \ref{lma:Q*} yields Lemma \ref{lma:model}.
\subsection{Boltzmann over Optimal Q: $\epB$}
\label{sec:boltz}

%zzz all this theory is quite abstract. is there some way to weave in a running example and/or to at least provide some more motiviation for why phiBoltz is desirable. Balancing exploration/exploitation isn't really a motivation for why values should be abstracted together. What do the first two abstraction types miss that this one provides? We didn't Lihong et al. study it in the exact case?
Here, we introduce the approximate abstraction family, $\epB$, which aggregates states with similar Boltzmann distributions on optimal $Q$-values as equivalent. This family of abstractions is appealing as the Boltzmman distribution naturally balances exploration and exploitation~\cite{sutton1998reinforcement}.

@@

By the approximation in Equation~\ref{eq:e_to_x_approx} and the assumption in Equation~\ref{eq:bolt_denom}:
\begin{align*}
\left|\frac{1 + Q_G(s_1,a) + \delta_1}{\sum_j e^{Q_G(s_1,a_j)}} - \frac{1 + Q_G(s_2,a) + \delta_2}{\sum_j e^{Q_G(s_1,a_j)} \pm k\epsilon}\right| \leq \epsilon
\end{align*}
It follows that:
\begin{equation}
\left|Q_G(s_1,a) - Q_G(s_2,a)\right| \leq \epsilon \left(\frac{|\mathcal{A}|}{1-\gamma} + k\epsilon + k \right).
\label{eq:bolt_qs}
\end{equation}
Consequently, we can consider $\epB$ as a variation of $\epQ$, where the difference in $Q$-values is bounded by $\epsilon \left(\frac{|\mathcal{A}|}{1-\gamma} + k\epsilon + k \right)$. Lemma~\ref{lma:bolt_lemma} then follows directly from Lemma~\ref{lma:Q*}.
